This chapter explored the most common static analyses used for strictness
analysis.  Because the initial placement of parallelism is crucial to our
technique we explored each possible analysis in depth, highlighting the
drawbacks of the two-point and four-point analyses. While projection-based
analysis is significantly more complex, its ability to determine the strictness
properties of functions based on the \emph{demand} placed on their results
provides too many benefits to ignore.

Many of the previous attempts and using strictness analysis for implicit
parallelism used the four-point analysis. Combined with \emph{evaluation
transformers} (discussed in the next section) the four-point analysis is able
to identify a significant amount of parallelism. Unfortunately, the analysis is
limited to functions on lists, which while ubiquitous, significantly restricts
the potential of identifying parallelism in more complex programs.

The projection-based analysis not only provides more insight into functions
like \<append\>, as discussed in Section \ref{sec:projSem}, but it also allows
us to determine the strictness properties for functions on arbitrary types.
This greatly expands the applicability of the strictness analysis for finding
potential parallelism.

The major drawback of the projection-based analysis is that we no longer know,
as compiler writers, what types of demand our analysis will return. With the
four-point analysis we can hard-code the parallel Strategies that correspond to
each of the points in the domain. If we expand our analysis to include pairs,
we can again add the corresponding strategies. With the projection-based
analysis we no longer have that foresight. This is the issue we address in the
next chapter.
