Less than a year ago I was going to quit my PhD. I had struggled throughout the
process and had decided that I was not cut out for it. It was not the first
time I almost quit, and it was not the first time that Prof. Runciman convinced
me to press on and get my work out into the world. I cannot imagine a better
supervisor. I feel grateful every day that he thought I
was worth keeping around.

In my six years in the UK I was almost deported once. That may sound bad, but
my partner, Danielle, \emph{was} deported. It is a testament to her level of
support for me that she went through the process of \emph{coming back} to the
country that deported her so that we would not have to live apart. It was not
easy, but it has been worth it.

My parents have never stopped me from following my curiosity. You would think
that after asking to go to music school and then switching to CS for grad
school they'd make me pick something and stick with it. I am so thankful to
have the supportive and understanding parents that I have.

Layla and Mike, for having the cutest baby. I can't wait to spend more time
with you three.

The members of York's PLASMA group have been a constant source of joy and
knowledge. I owe them many thanks. Matt Naylor, who patiently answered all my
silly questions during my first year. Jason Reich, who was always enthusiastic
about everyone's work. Chris Bak for empathising with my PhD progress and
playing board games with me. Glyn Faulkner for getting me excited about
different languages and technologies. And Rudy and Michael for humoring me
while I pitch them my half-baked language ideas.

Simon Poulding provided tons of insight into heuristic search and stats. Mike Dodds
for keeping his door open to me.

Ian, Gary, Jamie, and Gareth always made me feel welcome when I would intrude
on RTS breaks. Russell for understanding that sometimes I just want to watch TV
and vent. Jimmy for always hearing me out and his steadfast confidence in me.
Alan for his joke\footnote{An enormous elephant came up to me the other day,
looking confused.  He asked me: ``What year is this?''. I said ``2015''. He
said ``I must have travelled forwards in time 10,000 years!''. I said ``Yeah,
you Mast'o'don''.} Sam and Dani for always being such gracious hosts.

Rich, Chris, and Tara for helping me feel less homesick.

Danielle again, for putting up with who I am but always helping me be better.
